The Functional Therapeutic Chemical Classification System

### 2013-08-19 document.tex
First sections have been final proofed
Signed-off-by: loopasam <samuel.croset@gmail.com>

\section{Motivation:}
Drug repurposing is the discovery of new indications for compounds already approved and used in a clinical setting. 
Recently, some computational approaches have been developed to unveil new opportunities in a systematic fashion, by taking 
into consideration gene expression signatures or chemical features for instance. 
We present here a novel method based on knowledge integration using semantic technologies. 
In order to computationally generate repurposing hypotheses, we used the Web Ontology Language (OWL) to 
mathematically define the meaning of over 20'000 terms formally describing various modes and mechanisms of action.

\section{Results:}
Based on an integration of public data, we have automatically assigned over a thousand of approved drugs into 
these mode of action categories. The resulting new resource is called the Functional Therapeutic Chemical Classification System (FTC) 
and was further evaluated against the content of the traditional Anatomical Therapeutic Chemical Classification System (ATC). 
We illustrate how the new classification can be used to generate drug repurposing hypotheses, using Alzheimer’s disease as a use-case.

@@

interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, sildenafil was initially developed 

@@

A feature of particular interest to describe drugs is the mode of action (MoA). According to Wikipedia, 
the MoA describes a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance. For instance 
terms such as transcriptional regulation agent or anticoagulant define MoAs and characterise the roles of a certain type of drugs. The 

@@

In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions are manually curated in this resource 
and the meaning of terms is specified by the relations linking two GO terms.

@@

manually curated information. Moreover, the categories described in these classification systems are sometimes used to annotate 
drugs: For instance the compound sildenafil has been manually annotated as vasodilator agent (CHEBI:35620 or MeSH:D27.505.954.411.918).

@@

Most of the drugs are present in multiple FTC categories, reflecting the various roles a compound can play inside a biological system. 

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank \citep{Knox2011} provides the 
known links between drugs and their protein targets. Uniprot \citep{TheUniprotConsortium2013} maps targets to their 
respective GO annotations \citep{Dimmer2012}. 
Drugs are further assigned into mode of action categories based of the OWL constructs and axioms defined in the 
FTC (see Availability and Implementation). A reasoner, a program capable of understanding such axioms, performed this task. 
The process to build the FTC is summarised in Figure~\ref{fig:01}.

\begin{figure*}[!tpb]%figure1
\centerline{\includegraphics{fig1.png}}
\caption{Parent categories to the FTC class Pro-fibrinolysis agent (FTC\_P0042730). The classification is a direct 
acyclic graph where categories are describing increasingly specific concepts. Arrows entails subclass relationships 
between the terms (is a relation).}\label{fig:01}
\end{figure*}

It takes around 5 seconds on a modern 
laptop (4 cores processor with 4 Gb of memory) to classify the knowledge base using the ELK reasoner \citep{Kazakov2011}. 
Other OWL reasoners have been tried (e.g, Hermit, Pellet, etc.) but none of them were suitable for this purpose; 
the classification time was always superior to a few minutes (data not shown). The FTC has a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. Categories can have multiple parents and multiple children. 
The reader can browse and access the content of the FTC online at {{https://www.ebi.ac.uk/chembl/ftc}}.

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics{fig2.png}}
\caption{Parent categories to the FTC class Pro-fibrinolysis agent (FTC\_P0042730). The classification is a direct 
acyclic graph where categories are describing increasingly specific concepts. Arrows entails subclass relationships 
between the terms (is a relation).}\label{fig:02}
\end{figure}

In total there are 1280 approved DrugBank compounds (chemical and biotherapeutics) perturbing 1264 human protein targets. 
The FTC introduces 23'353 new categories describing the mode and mechanism of action of therapeutic compounds. 4289 of these 
categories are related to biological processes and 19'064 to the molecular functions. A summary of the metrics behind the latest 
build is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}. Out of all FTC classes, 1432 
categories (\textgreater 6\%) directly contain at least one approved drug. This number increases up to 2532 (\textgreater 11\%) 
when direct and 
indirect drugs are considered. FTC categories not containing drugs (e.g, FTC\_A0001771 - Anti-immunological synapse 
formation agent) represent modes and mechanisms of action for which no approved compounds exist yet or that have not been identified 
as such in the FTC.

@@

\section{Availability and Implementation}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the FTC using the Brain library \citep{Croset2013}, and can be 
find at {{https://www.ebi.ac.uk/chembl/ftc}}. The documentation can be accessed at {{https://github.com/loopasam/ftc/wiki}}. The full
description of the methodology used to generate the classification is available as supplementary material on Bioinformatics website.

>>>>

\section{Motivation:}
Drug repurposing is the discovery of new indications for compounds already approved and used in a clinical setting. 
Recently, some computational approaches have been developed to unveil new opportunities in a systematic fashion, by taking 
into consideration gene expression signatures or chemical features for instance. 
We present here a novel method based on knowledge integration using semantic technologies, 
to capture the functional role of approuved chemical compounds.
\section{Results:}
In order to computationally generate repurposing hypotheses, we used the Web Ontology Language (OWL) to 
mathematically define the meaning of over 20'000 terms formally describing various modes and mechanisms of action. Based on an 
integration of public data, we have automatically assigned over a thousand of approved drugs into 
these mode of action categories. The resulting new resource is called the Functional Therapeutic Chemical Classification System (FTC) 
and was further evaluated against the content of the traditional Anatomical Therapeutic Chemical Classification System (ATC). 
We illustrate how the new classification can be used to generate drug repurposing hypotheses, using Alzheimer’s disease as a use-case.

@@

interesting repurposing opportunities \citep{Medina-Franco2013}. For instance, \emph{sildenafil} was initially developed 

@@

A feature of particular interest to describe drugs is the Mode of Action (MoA). According to Wikipedia, 
the MoA describes \emph{a functional or anatomical change, 
at the cellular level, resulting from the exposure of a living organism to a substance}. For instance 
terms such as \emph{transcriptional regulation agent} or \emph{anticoagulant} define MoAs and characterise the 
roles of a certain type of drugs. The 

@@

In the Gene Ontology (GO) for example \citep{Ashburner2000}, biological processes and molecular functions terms are manually curated  
and their meaning is specified by the relation types linking two GO terms.

@@

manually curated information; moreover, the categories described in these classification systems are sometimes used to annotate 
drugs: For instance the compound \emph{sildenafil} has been manually annotated as \emph{vasodilator agent} (CHEBI:35620 or MeSH:D27.505.954.411.918).

@@

Most of the drugs are present in multiple FTC categories, reflecting the various roles a compound can play inside a biological system which 
can serve as starting point for repurposing. 

@@

The knowledge base behind the FTC is built by integrating information coming from various sources. 
The GO terms serve as template to create the FTC categories describing the MoAs; DrugBank \citep{Knox2011} provides the 
known links between drugs and their protein targets and Uniprot \citep{TheUniprotConsortium2013} maps targets to their 
respective GO annotations \citep{Dimmer2012}. 
Drugs are further assigned into MoA categories based of the OWL constructs and axioms defined in the 
FTC (see \emph{Availability and Implementation}). A reasoner, a program capable of understanding such axioms, performed this task. 
The process to build the FTC is summarised in Figure~\ref{fig:01}.
It takes around 5 seconds on a modern 
laptop (4 cores processor with 4 Gb of memory) to classify the knowledge base using the ELK reasoner \citep{Kazakov2011}. 
Other OWL reasoners have been tried (e.g, Hermit, Pellet, etc.) but none of them were suitable for this purpose; 
the classification time was always superior to a few minutes (data not shown). The FTC has a taxonomic structure as illustrated 
on Figure~\ref{fig:02}, which arises when the reasoner classifies the knowledge base. Categories can have multiple parents and multiple children. 
The reader can browse and access the content of the FTC online at {{https://www.ebi.ac.uk/chembl/ftc}}.
In total there are 1280 approved DrugBank compounds (chemical and biotherapeutics) perturbing 1264 human protein targets. 
The FTC introduces 23'353 new categories describing the mode and mechanism of action of therapeutic compounds. 4289 of these 
categories are related to biological processes and 19'064 to the molecular functions. A summary of the metrics behind the latest 
build is available online at {{https://www.ebi.ac.uk/chembl/ftc/evaluation/}}. Out of all FTC classes, 1432 
categories (\textgreater 6\%) directly contain at least one approved drug. This number increases up to 2532 (\textgreater 11\%) 
when direct and 
indirect drugs are considered. FTC categories not containing drugs (e.g, FTC\_A0001771 - \emph{Anti-immunological synapse 
formation agent}) represent modes and mechanisms of action for which no approved compounds exist yet or that have not been identified 
as such in the FTC.

\begin{figure*}[!tpb]%figure1
\centerline{\includegraphics{fig1.png}}
\caption{Summary of the methodology used to construct the resource. The FTC categories are first created (A) from the existing GO terms
describing molecular functions and biological processes. Then, each category has an equivalent definition assign to it (B).
The content of various databases in incorporated (C) and linked using the FTC specific properties. Finally a reasoner runs over the knowledge
base and assigns drugs to MoA categories based on whether or not a definition can be satisfied.}\label{fig:01}
\end{figure*}

\begin{figure}[!tpb]%figure2
\centerline{\includegraphics{fig2.png}}
\caption{Parent categories to the FTC class \emph{Pro-fibrinolysis agent} (FTC\_P0042730). The classification is a direct 
acyclic graph where categories are describing increasingly specific concepts. Arrows entails subclass relationships 
between the terms (\emph{is a} relation).}\label{fig:02}
\end{figure}

@@

\section{Availability and Implementation}\label{impl}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the FTC using the Brain library \citep{Croset2013}, and can be 
find at {{https://www.ebi.ac.uk/chembl/ftc}}. The documentation can be accessed at {{https://github.com/loopasam/ftc/wiki}}. The full
description of the methodology used to generate the classification is available as supplementary material on Bioinformatics website.
